# Example with fish data from NORS

In this example we are interested in exploring data from a specific data resource -- Swedish Electrofishing Registry - SERS (Institutionen för akvatiska resurser, SLU). This data base has 2.8 M observations starting in the 1950's.

As you may already know, SBDI is a collection of many biodiversity databases. We start by searching for the data resource we are interested in using the function `pick_filter()`. This is an interactive query guiding you through the many resources available to filtering your query (data resources, spatial layers, and curated species lists).

```{r filter1, message=FALSE, eval=FALSE}
library(SBDI4R)
fq_str <- pick_filter("resource") 
# follow the instructions 
```

Follow the instruction. Your choices here would have been "in3" --> "dr10". Your variable `fq_str` will now contain a string "data_resource_uid:dr10".

```{r filter2}
# we cant do it interactive here so we force it
fq_str <- "data_resource_uid:dr10"
```

But we are not interested in the complete database, but on the last 10 years of data. for this we concatenate (add to a vector) another filter string. These will be treated as AND factors.

```{r filter3}
y1 <- 2012
y2 <- 2021
fq_str <- c(fq_str, paste0("year:[", y1, " TO ", y2,"]"))
# Note the square brackets are hard limits
```

For references on how to use the filters see SBDI APIS [documentation](https://api.biodiversitydata.se/?lang=en-US#ws3).

Using the function `occurrences()` we can the query for the observations fulfilling our filter. If you haven't specified that in the `sbdi_config()` before, you need to pass your email and the download reason.

```{r getOcc, cache=FALSE}
library(SBDI4R)
xf <- occurrences(fq = fq_str,
                 email = "alejandro@greensway.se", 
                 download_reason_id=10)

# Simply summarise all records by data source 
table(xf$data$dataResourceName)
table(xf$data$dataResourceID)
```

## Plotting data on a map

You can quickly plot all the observations as a PDF file with the function `ocurrence_plot()`, one page per species:

```{r pdf_plot, eval=FALSE}
occurrences_plot(xf, "obsPlot.pdf", 
                 grouped=FALSE, 
                 taxon_level="species", 
                 pch='+')
```

Note that the plot is saved to a pdf file in the current working directory. You can find that with `getwd()`.

#### Leaflet

There are many other ways of producing spatial plots in R. The leaflet package provides a simple method of producing browser-based maps with panning, zooming, and background layers:

```{r leaflet, message=FALSE, warning=FALSE}
library(leaflet)
# drop any records with missing lat/lon values
xfl <- xf$data[!is.na(xf$data$longitude) | !is.na(xf$data$latitude),] 
marker_colour <- rep("#d95f02", nrow(xfl))
# blank map, with imagery background
leaflet(width = "100%") %>% 
  addProviderTiles("Esri.WorldImagery") %>%
  # add markers
  addCircleMarkers(xfl$longitude, xfl$latitude,  
                   radius = 1, 
                   fillOpacity =.5, 
                   opacity = 1,
                   col=marker_colour,
                  clusterOptions = markerClusterOptions())
```

## Temporal summary

A quick summary over the years reveal a drop in number of records over time <span class="hgl>[IF DATA IS NOT CHANGED]</span>.

```{r timeTable}
table(xf$data$year)
```

```{r timeHist}
hist(xf$data$year, 
     breaks = seq(y1, y2), 
     xlab = "Year", 
     main = "")
```

## Species summary

In the same way we can summaries the number of observations for each species, by common or scientific name.
```{r nObsTab, paged.print=TRUE}
sppTab <- table(xf$data$commonName)
sppDF <- as.data.frame(sppTab)
colnames(sppDF)[1] <- "species"
sppDF
```

```{r nObsTab2, paged.print=TRUE}
sppTab <- table(xf$data$scientificName)
sppDF <- as.data.frame(sppTab)
colnames(sppDF)[1] <- "species"
sppDF
```

Perhaps, you need to send this table as a .CSV file to a colleague.

```{r savesummary, eval=FALSE}
write.csv(sppDF, "NORS species summary.csv")
# NOTE: again this will be saved on your working directory
```

## Spatial biodiversity analysis

Let's now ask: how does the species richness vary across Sweden? In this case we want to summarise occurrences species-wise over a defined grid instead of plotting every observation point. First we need to overlay the observations with a grid. In this case, the standard Swedish grids at 50, 25, 10 and 5 km are provided as data in the SBDI4R package (with Coordinate Reference System = WGS84, EPSG:4326).

```{r overlay}
library(sp) # the function coordinates() and proj4string() are in sp
library(rgeos) # the function over() is in package rgeos
# load some shapes over Sweden's political borders
data("swe_wgs84", package="SBDI4R", envir=environment())
# A standard 50km grid
data("Sweden_Grid_50km_Wgs84", package="SBDI4R", envir=environment())

grid <- Sweden_Grid_50km_Wgs84

# make the observations spatial
# NOTE: make sure there are no NAs on either column defining the coordinates
# xf$data[!is.na(xf$data$longitude) | !is.na(xf$data$latitude),]

obs <- as.data.frame(xf$data)
coordinates(obs) <- obs[,c("longitude","latitude")]
wkt <- sf::st_crs(4326)[[2]]
proj4string(obs) <- sp::CRS(wkt) #CRS("+init=epsg:4326")

nObs <- nrow(obs)

# overlay the data with the grid
ObsInGridList <- over(grid, obs, returnList=TRUE)
wNonEmpty <- unname( which( unlist(lapply(ObsInGridList, nrow)) != 0) )
if(length(wNonEmpty)==0) message("Observations don't overlap any grid cell.")
```

The result `ObsInGridList` is a `list` object with a subset of the data on each grid. Now summarise occurrences within grid cells:

```{r overlay_summary}
# check n the total number of observations
sum(unlist(lapply(ObsInGridList, nrow)))

# apply a summary over the grid cells 
nCells <- length(ObsInGridList)

res <- data.frame("nObs"=as.numeric(rep(NA,nCells)),
                  "nYears"=as.numeric(rep(NA,nCells)),
                  "nSpp"=as.numeric(rep(NA,nCells)),
                  row.names = row.names(grid),
                  stringsAsFactors = FALSE)

cols2use <- c("scientificName", "year")

dataRes <- lapply(ObsInGridList[wNonEmpty], 
                  function(x){
                    x <- x[,cols2use]
                    colnames(x) <- c("scientificName", "year")
                    return(c("nObs" = length(x[,"scientificName"]),
                             "nYears" = length(unique(x[,"year"])),
                             "nSpp" = length(unique(x[,"scientificName"]))
                             )
                           )
                    }
                  )

dataRes <- as.data.frame(dplyr::bind_rows(dataRes, .id = "gridID"))

res[wNonEmpty,] <- dataRes[,-1]

resSp <- sp::SpatialPolygonsDataFrame(grid, res)
```

And finally plot the grid summary as a map:
```{r grid, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
palBW <- leaflet::colorNumeric(c("white", "navyblue"),
                               c(0, max(resSp@data$nSpp, na.rm = TRUE)),
                               na.color = "transparent")
oldpar <- par()
par(mar = c(1,1,0,0))
plot(resSp, col=palBW(resSp@data$nSpp), border = NA)
plot(swe_wgs84$Border, border=1, lwd=1, add=T)
legend("bottomleft", 
       legend = round(seq(0, max(resSp@data$nSpp, na.rm = TRUE), length.out = 5)),
       col = palBW(seq(0, max(resSp@data$nSpp, na.rm = TRUE), length.out = 5)),
       title = "Number of \nspecies", pch = 15, bty="n")
suppressWarnings(par(oldpar))
```

We can go further by gathering the observations by latitude. 
[TO BE CONTINUED]
```{r sppLatitude}
library(dplyr)
library(tidyr)
xgridded <- xf$data %>%
    ## discard genus- and higher-level records
    filter(rank %in% c("species", "subspecies", "variety", "form", "cultivar")) %>%
    mutate(longitude = round(longitude * 4)/4, 
           latitude = round(latitude * 4)/4) %>%
    group_by(longitude,latitude) %>%
    ## subset to vars of interest
    select(longitude, latitude, species) %>%
    ## take one row per cell per species (presence)
    distinct() %>%
    ## calculate species richness
    mutate(richness=n()) %>%
    ## convert to wide format (sites by species)
    mutate(present=1) %>%
    do(tidyr::pivot_wider(data=.,  names_from=species, values_from=present, values_fill=0)) %>%
    ungroup()
## where a species was not present, it will have NA: convert these to 0
sppcols <- setdiff(names(xgridded),
                   c("longitude", "latitude", "richness"))
xgridded <- xgridded %>% 
  mutate_at(sppcols, function(z) ifelse(is.na(z), 0, z))
```

And plot it accordingly
```{r plot_richLat, warning=FALSE}
library(ggplot2)

ggplot(xgridded, aes(latitude, richness)) + 
  labs(x = "Latitude (º)", 
       y = "Species richness") +
  lims(y = c(0,20)) +
  geom_point() + 
  theme_bw()
```


## Temporal biodiversity analysis
Despite the drop seen in sampling effort <span class=hgl>(IF DATA IS NOT CHANGED)</span>, we can still analyse weather there is any trends for species over the 8 yrs?:

```{r}
xf$data %>%
    ## discard genus- and higher-level records
    filter(rank %in% c("species", "subspecies", "variety", "form", "cultivar")) %>% 
    group_by(year) %>%
    ## subset to vars of interest
    select(year, species) %>%
    ## take one row per cell per species (presence)
    distinct() %>%
    ## calculate species richness
    mutate(richness=n()) %>%
    ## convert to wide format (sites by species)
    mutate(present=1) %>%
    do(tidyr::pivot_wider(data=.,  names_from=species, values_from=present, values_fill=0)) %>%
    ungroup()
  
```

First we check whether spatial sampling effort was comparable across those years – can we quickly check number of sampling sites over years?
And we ask: can we see a trend (change = frequency of occurrence = number caught) for species X, or species Y?

## Links to further tutorials that could be of interest
Link to analyses of interest – e.g. species distribution models, or trend analyses, or species divesity – link to biodiversity workflow

