# Example with opportunistic data on Dragonflies

In this example we are interested in exploring opportunistically collected data 
from the Swedish citizen science species observation portal - Artportalen.

## Name searching
To begin with, we want be sure there is an unequivocal way to find the species within 
the order Odonata (dragonflies) and nothing else, so let's search for "odonata":
```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE, include=FALSE}
library(SBDI4R)
```

```{r search1, echo=TRUE, eval=FALSE}
sx <- search_fulltext("odonata")
sx$data[, c("guid", "scientificName", "rank", "occurrenceCount")]
```  

```{r search1eval, echo=FALSE}
sx <- search_fulltext("odonata")
sx$data[, c("guid", "scientificName", "rank", "occurrenceCount")]
```  

We quickly see there that other taxonomic levels appear too, and also species that look suspiciously as not belonging to 
dragonsflies. But there is only one order. Let's refine the search. To know which search fields we can use 
to filter the search we use the function `sbdi_fields(fields_type = "general")`.
The search field we are looking for is "order_s".  
```{r search2,  echo=TRUE, eval=FALSE}
sx <- search_fulltext(fq="order_s:Odonata", page_size = 10)
sx$data[, c("scientificName", "rank", "occurrenceCount")]
```  
```{r search2eval, echo=FALSE}
sx <- search_fulltext(fq="order_s:Odonata", page_size = 10)
sx$data[, c("guid", "scientificName", "rank", "occurrenceCount")]
```  
Now we can download the taxonomic data (note that the search is case-sensitive):
```{r, message=FALSE, results=FALSE}
tx <- taxinfo_download("order_s:Odonata", 
                       fields = c("guid", "order_s","genus_s", "specificEpithet_s", 
                                  "scientificName",  "canonicalName_s", "rank"), 
                       verbose = FALSE)
tx <- tx[tx$rank == "species" & tx$genusS != "",] ## restrict to species and not hybrids
```
You can save the `tx` object as the complete species list for later use. 

## Filter the search to get the observations
We start by searching for the data resource we are interested in using 
the function `pick_filter()`. This is an interactive query guiding you through 
the many resources available to filtering your query (data resources, spatial 
layers, and curated species lists).

```{r filter2.1, message=FALSE, eval=FALSE}
# follow the instructions 
fq_str <- pick_filter("resource") 
``` 

Follow the instructions. Your choices here would have been "in3" --> "dr5". 
Your variable `fq_str` will now contain a string "data_resource_uid:dr5".

```{r filter2.2, echo=FALSE, eval=TRUE}
# we cant do it interactive here so we force it
fq_str <- "data_resource_uid:dr5"
```

We only want to look at data from year 2000 to 2010:
```{r filter2.3}
y1 <- 2000
y2 <- 2010
fq_str <- c(fq_str, paste0("year:[", y1, " TO ", y2,"]"))
# Note the square brackets are hard limits
```

We also want to filter spatially for Southern Sweden ([Götaland](https://en.wikipedia.org/wiki/G%C3%B6taland)).

Vector spatial layers (eg. polygons) can be imported  in a number of different ways. 
SBDI APIs take as search input polygons in the so-called WKT [Well Known Text](https://www.geoapi.org/3.0/javadoc/org/opengis/referencing/doc-files/WKT.html) 
format. So the first step is to load a vector layer and transform it into a WKT string. 
You could instead use the data we provid in the SBDI4R package `data("swe")`. 

```{r sweshape}
data("swe",package = "SBDI4R")

wGotaland <- swe$Counties$LnNamn %in% c("Blekinge", "Gotlands", "Hallands", 
                                        "Jönköpings", "Kalmar", "Kronobergs", 
                                        "Östergötlands", "Skåne", "Västra Götalands")
gotaland_c <- swe$Counties[wGotaland,]
```

There are details about this polygon that we need to take care before. The WKT 
string should not be too long to be accepted by the API service. Also, the polygon
we just got is projected in the coordinate system SWEREF99 TM, and the API service 
only accepts coordinates in a geodesic coordinate system WGS84. Let's construct 
the WKT string:
```{r searchpoly2WKT, warning=FALSE}
# transform the CRS
gotaland_c <- sf::as_Spatial(
                sf::st_transform(
                  sf::st_as_sf(gotaland_c), 
                  crs = sf::st_crs(4326)$wkt) )

# disolve the counties into one polygon
gotaland <- rgeos::gUnaryUnion(gotaland_c)

# extract the polygons coordinates
nPol <- length(gotaland@polygons[[1]]@Polygons) 
lonlat <- list()
for(p in seq(nPol)){
  lonlat[[p]] <- gotaland@polygons[[1]]@Polygons[[p]]@coords
}
lonlat <- do.call(rbind, lonlat)

# create a convex hull of the polygon to simplify the geometry and 
# reduce the length of the WKT string
gotaland_ch <- chull(lonlat)
lonlat <- lonlat[c(gotaland_ch, gotaland_ch[1]), ]

# create WKT string
# first join each lon-lat coordinate pair
wkt_temp <- apply(lonlat, 1, function(z) paste(round(z,4), collapse=" "))
# now build the WKT string
wkt <- paste("MULTIPOLYGON(((", paste(wkt_temp, collapse=","), ")))", sep="")
# NOTE: as of today, the SBDI APIs will only work properly if the polygon is 
# submitted as a MULTIPOLYGON
```
The WKT string then looks like this:
```{r}
wkt
```
```{r searchpoly, fig.width=7, echo=FALSE, warning=FALSE}
library(sp)
oldpar <- par()
par(mar = c(0,0,0,0))
plot(gotaland_c, border="blue")
plot(gotaland, lwd=2, add=TRUE)
plot(rgeos::readWKT(wkt), lwd=2, border="red", add=TRUE)
legend("bottomleft", inset = c(0,0.05),
       legend = c("counties", "Götaland", "search polygon"),
       col = c("blue","black", "red"),
       lwd = c(1,2,1),
       title = "Polygons",bty="n")
suppressWarnings(par(oldpar))
```

Next, we download the observations using the command `occurrences()`, but be aware that 
the search fields may not be the same as those used to search for taxa. We therefore
recommend using the function `sbdi_fields("occurrence")` to find out which search fields we can use 
to filter for occurrences. Here we see that the field 
we need this time is "order".

```{r getData, cache=TRUE}
xf <- SBDI4R::occurrences(taxon = "order:Odonata", 
                  fq = fq_str,
                  wkt = wkt,
                  extra = "collector",
                  email = "sbdi4r-test@biodiversitydata.se", 
                  download_reason_id = 10)
```
```{r localtmp, echo=FALSE, eval=FALSE}
save(xf, file = "data/tmp_occdata2_for compile.rdata")
load(file = "data/tmp_occdata2_for compile.rdata")
```
We have now downloaded the data locally and depending on your configuration this
will be cached on your computer. However, as the search and download could take
long time, we recommend to save the data locally. 
```{r save2, eval=FALSE}
save(xf, file = "an_approprieted_name.rdata")
load(file = "an_approprieted_name.rdata")
```
## Quality and fit-for-use check
Before we can use the observation records we need to know if the observation 
effort (sampling effort) has varied over time and in space. We can approximate observation effort from the data by 
defining field visits i.e. occasions at which an observer has sampled observations. We reconstruct field visits
(that is, assign each observation a visitUID) using using the package 
[BIRDS](https://greensway.github.io/BIRDS/). Additionally we want the data to be summarized 
over a grid of 25 km (provided through the SBDI4R package). The following functions 
will perform many different summaries at the same time. Please refer to the BIRDS package 
documentation for more detail. 

```{r birds}
library(BIRDS)
OB <- organiseBirds(xf$data, sppCol = "species" , 
                    # We only want observations identified at the species level
                    taxonRankCol = "rank", taxonRank = "species", 
                    # the visits are defined by collector and named locality
                    idCols = c("locality", "collector"), 
                    timeCols = c("year", "month", "day"), 
                    xyCols =c("longitude","latitude") )

# We don't need the whole grid, just the piece that overlaps our searching polygon
gotaland_grid25 <- raster::intersect(gotaland, Sweden_Grid_25km_Wgs84)

# This is another way of doing it.
# gotaland_grid25 <- gIntersection(gotaland,
#                                  spTransform(Sweden_Grid_25km_Wgs84, 
#                                              CRSobj = CRS(sf::st_crs(4326)$wkt)))

SB <- summariseBirds(OB, grid = gotaland_grid25, spillOver = "unique")
```

Once summarised, we can see over space and for a few selected years how the 
number of observations is distributed:
```{r plotBIRDSspatial_cose, eval=FALSE}
maxC <- max(SB$spatial@data$nObs, na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")
oldpar <- par()
par(mar = c(4,0,4,0), mfrow=c(1,3))
plot(SB$spatial, col=palBW(SB$spatial@data$nObs),
     border = "grey", main="All years") ## with palette
legend("bottomleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       title = "Number of \nobservations", pch = 15, bty="n")

## or export other combinations, e.g. one map per observed year
yearlySp <- exportBirds(SB, 
                        dimension = "spatial", 
                        timeRes = "yearly", 
                        variable = "nObs", 
                        method = "sum")

maxC <- max(yearlySp@data$'2005', na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")

plot(yearlySp["2005"], col=palBW(yearlySp@data$'2005'), 
     border = "grey",main="2005")
legend("bottomleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       border = "grey",
       title = "Number of \nobservations", pch = 15, bty="n")

maxC <- max(yearlySp@data$'2010', na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")

plot(yearlySp["2010"], col=palBW(yearlySp@data$'2010'), 
     border = "grey",main="2010")
legend("bottomleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       border = "grey",
       title = "Number of \nobservations", pch = 15, bty="n")
par(oldpar)
```
```{r plotBIRDSspatial, echo=FALSE, fig.width=7}
maxC <- max(SB$spatial@data$nObs, na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")
oldpar <- par()
par(mar = c(4,0,4,0), mfrow=c(1,3))
plot(SB$spatial, col=palBW(SB$spatial@data$nObs),
     border = "grey", main="All years") ## with palette
legend("topleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       title = "Number of \nobservations", pch = 15, bty="n")

## or export other combinations, e.g. one map per observed year
yearlySp <- exportBirds(SB, 
                        dimension = "spatial", 
                        timeRes = "yearly", 
                        variable = "nObs", 
                        method = "sum")

maxC <- max(yearlySp@data$'2005', na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")

plot(yearlySp["2005"], col=palBW(yearlySp@data$'2005'), 
     border = "grey",main="2005")
legend("topleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       border = "grey",
       title = "Number of \nobservations", pch = 15, bty="n")

maxC <- max(yearlySp@data$'2010', na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")

plot(yearlySp["2010"], col=palBW(yearlySp@data$'2010'), 
     border = "grey",main="2010")
legend("topleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       border = "grey",
       title = "Number of \nobservations", pch = 15, bty="n")
suppressWarnings(par(oldpar))
```

We now want to use the number of field visits as the measure
for sampling effort. And since there are other ways to plot spatial data, we this time
use the package `sf` instead of `sp`:  
```{r ggplot1, message=FALSE, warning=FALSE, fig.width=7}
library(sf)
library(cowplot)
library(ggplot2)
library(colorRamps)
library(gridExtra)

spatial_sf <- st_as_sf(SB$spatial)

vis <- ggplot(data = spatial_sf, aes( fill = nVis)) +
  geom_sf() +
  ggtitle("Visits") +
  scale_fill_gradient(low = "#56B1F7",
                      high = "#132B43",
                      na.value = NA) +
  theme(plot.margin = margin(1, 1, 1, 1, "pt")) +
  theme_cowplot()

spp <- ggplot(data = spatial_sf ,aes( fill = nSpp))+
  geom_sf()+
  ggtitle("Number of species")+
  scale_fill_gradient(low = "#56B1F7",
                      high = "#132B43",
                      na.value = NA) +
  theme(plot.margin = margin(1, 1, 1, 1, "pt")) +
  theme_cowplot()

grid.arrange(vis, spp, ncol=2)
```

We see that `SB` contains an element called `SB$temporal` that contains a daily 
time series with time-specific rows when there is information. `xts` also supports 
day time, but dating below day resolution is not yet implemented in the `BIRDS` package.  

```{r temporal}
sb.xts <- SB$temporal
dim(sb.xts)
head(sb.xts, 20)
```

Sub-setting is convenient in `xts` as you can do it with its dates and with a `/` 
for a range of dates.  
```{r subsetting, message=FALSE, warning=FALSE}
sb.xts["2010-09-07"] #a specific day
sb.xts["2010-09-01/2010-09-15"] #for a period
sb.xts["2010-09"] #a specific month
```

The package `xts` has several tools for converting to different time periods. Here we 
will use `to.monthly`. This provides, the first, min, max, and last of the data. 
We can plot the daily maximum number of observations. The plot command with an 
`xts` object provides a TON of features. This makes it fairly easy to customize 
your plots. Read more in `?plot.xts`.  

```{r monthlyPlotcode, eval=FALSE}
library(xts)
obs.m <- to.monthly(sb.xts$nObs) 
plot(obs.m["2000/2010",2], 
     col = "darkblue", 
     grid.ticks.on = "month", 
     major.ticks = "month", 
     grid.col = "lightgrey",  
     main = "Maximum number of daily observations/visits per month")

vis.m <- to.monthly(sb.xts$nVis) 
lines(vis.m["2000/2010",2], col = "orange", lwd=2)
```

```{r monthlyPlot, echo=FALSE, fig.height=5, fig.width=9}
library(xts)
obs.m <- to.monthly(sb.xts$nObs) 
plot(obs.m["2000/2010",2], 
     col = "darkblue", 
     grid.ticks.on = "month", 
     major.ticks = "month", 
     grid.col = "lightgrey",  
     main = "Maximum number of daily observations and visits per month")

vis.m <- to.monthly(sb.xts$nVis) 
lines(vis.m["2000/2010",2], col = "orange", lwd=2)
```


We can now look at some particular species and ask whether this has changed in occurrence over time:
Plot no. records of species x and no. visits all species over years (we simply explore by comparing records for a species with no visits, can assume that species has increased of stronger positive trend than for no. visits)

Plot no. gridcells with visits for species x and no. gridcells with visits for all species over years (we simply explore by comparing records for a species with no visits, can assume that species has increased of stronger positive trend than for no. visits)
(species x: Tvåfläckad trollslända Epitheca bimaculata)

