---
output: html_document
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

# Example with opportunistic data on Dragonflies

In this example we are interested in exploring opportunistically
collected data from the Swedish citizen science species observation
portal - Artportalen.

## Name searching

To begin with, we want be sure there is an unequivocal way to find the
species within the order Odonata (dragonflies) and nothing else, so
let's search for "odonata":

```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE, include=FALSE}
library(SBDI4R)
library(sf)
library(BIRDS)
```

```{r search1, echo=TRUE, eval=FALSE}
sx <- search_fulltext("odonata")
sx$data[, c("guid", "scientificName", "rank", "occurrenceCount")]
```

```{r search1eval, echo=FALSE}
sx <- search_fulltext("odonata")
sx$data[, c("guid", "scientificName", "rank", "occurrenceCount")]
```

We quickly see there that other taxonomic levels appear too, and also
species that look suspiciously as not belonging to dragonflies. But
there is only one order. Let's refine the search. To know which search
fields we can use to filter the search we use the function
`sbdi_fields(fields_type = "general")`. The search field we are looking
for is "order_s".

```{r search2,  echo=TRUE, eval=FALSE}
sx <- search_fulltext(fq = "order_s:Odonata", page_size = 10)
sx$data[, c("scientificName", "rank", "occurrenceCount")]
```

```{r search2eval, echo=FALSE}
sx <- search_fulltext(fq = "order_s:Odonata", page_size = 10)
sx$data[, c("guid", "scientificName", "rank", "occurrenceCount")]
```

Now we can download the taxonomic data (note that the search is
case-sensitive):

```{r, message=FALSE, results=FALSE}
tx <- taxinfo_download("order_s:Odonata", 
                       fields = c("guid", "order_s","genus_s", "specificEpithet_s", 
                                  "scientificName",  "canonicalName_s", "rank"), 
                       verbose = FALSE)
tx <- tx[tx$rank == "species" & tx$genusS != "",] ## restrict to species and not hybrids
```

You can save the `tx` object as the complete species list for later use.

## Filter the search to get the observations

We start by searching for the data resource we are interested in using
the function `pick_filter()`. This is an interactive query guiding you
through the many resources available to filtering your query (data
resources, spatial layers, and curated species lists).

```{r filter2.1, message=FALSE, eval=FALSE}
# follow the instructions 
fq_str <- pick_filter("resource") 
```

Follow the instructions. Your choices here would have been "in3" --\>
"dr5". Your variable `fq_str` will now contain a string
"data_resource_uid:dr5".

```{r filter2.2, echo=FALSE, eval=TRUE}
# we cant do it interactive here so we force it
fq_str <- "data_resource_uid:dr5"
```

We only want to look at data from year 2000 to 2010:

```{r filter2.3}
y1 <- 2000
y2 <- 2010
fq_str <- c(fq_str, paste0("year:[", y1, " TO ", y2,"]"))
# Note the square brackets are hard limits
```

We also want to filter spatially for Southern Sweden
([Götaland](https://en.wikipedia.org/wiki/G%C3%B6taland)).

Vector spatial layers (eg. polygons) can be imported in a number of
different ways. SBDI APIs take as search input polygons in the so-called
WKT [Well Known
Text](https://www.geoapi.org/3.0/javadoc/org/opengis/referencing/doc-files/WKT.html)
format. So the first step is to load a vector layer and transform it
into a WKT string. You could instead use the data we provid in the
SBDI4R package `data("swe")`.

```{r sweshape}
data("swe",package = "SBDI4R")
wGotaland <- swe$Counties$LnNamn %in% c("Blekinge", "Gotlands", "Hallands", 
                                        "Jönköpings", "Kalmar", "Kronobergs", 
                                        "Östergötlands", "Skåne", "Västra Götalands")
gotaland_c <- swe$Counties[wGotaland,]
```

There are details about this polygon that we need to take care before.
The WKT string should not be too long to be accepted by the API service.
Also, the polygon we just got is projected in the coordinate system
SWEREF99 TM, and the API service only accepts coordinates in a geodesic
coordinate system WGS84. Let's construct the WKT string:

```{r searchpoly2WKT, warning=FALSE}
# transform the CRS
gotaland_c <- st_transform(gotaland_c,
                           crs = st_crs(4326))

# disolve the counties into one polygon
gotaland <- st_union(gotaland_c)

# create a convex hull of the polygon to simplify the geometry and 
# reduce the length of the WKT string
gotaland_ch <- st_convex_hull(gotaland)

# cast it as MULTIPOLYGON as this is what SBDIs API need
# NOTE: as of today, the SBDI APIs will only work properly if the polygon is 
# submitted as a MULTIPOLYGON
gotaland_ch <- st_cast(gotaland_ch, to = "MULTIPOLYGON")

# create WKT string
wkt <- st_as_text(gotaland_ch)

```

The WKT string then looks like this:

```{r, echo=FALSE}
wkt
```

```{r searchpoly, fig.width=7, echo=FALSE, warning=FALSE}
#library(sf)
oldpar <- par()
par(mar = c(0,0,0,0))
plot(gotaland_c$geometry, border = "blue")
plot(gotaland, lwd = 2, add = TRUE)
plot(rgeos::readWKT(wkt), lwd = 2, border = "red", add = TRUE)
legend("bottomleft", inset = c(0,0.05),
       legend = c("counties", "Götaland", "search polygon"),
       col = c("blue","black", "red"),
       lwd = c(1,2,1),
       title = NULL,bty = "n")
suppressWarnings(par(oldpar))
```

Next, we download the observations using the command `occurrences()`,
but be aware that the search fields may not be the same as those used to
search for taxa. We therefore recommend using the function
`sbdi_fields("occurrence")` to find out which search fields we can use
to filter for occurrences. Here we see that the field we need this time
is "order".

```{r getData, cache=TRUE}
xf <- occurrences(taxon = "order:Odonata", 
                  fq = fq_str,
                  wkt = wkt,
                  extra = "collector",
                  email = "sbdi4r-test@biodiversitydata.se", 
                  download_reason_id = 10)
```

```{r localtmp, echo=FALSE, eval=FALSE}
save(xf, file = "data/tmp_occdata2_for compile.rdata")
load(file = "data/tmp_occdata2_for compile.rdata")
```

We have now downloaded the data locally and depending on your
configuration this will be cached on your computer. However, as the
search and download could take long time, we recommend to save the data
locally. appropriate

```{r save2, eval=FALSE}
save(xf, file = "an_appropriate_name.rdata")
load(file = "an_appropriate_name.rdata")
```

## Quality and fit-for-use check

Before we can use the observation records we need to know if the
observation effort (sampling effort) has varied over time and in space.
We can approximate observation effort from the data by defining field
visits i.e. occasions at which an observer has sampled observations. We
reconstruct field visits (that is, assign each observation a visitUID)
using using the package [BIRDS](https://greensway.github.io/BIRDS/).
Additionally we want the data to be summarized over a grid of 25 km
(provided through the SBDI4R package). The following functions will
perform many different summaries at the same time. Please refer to the
BIRDS package documentation for more detail.

```{r birds, message=FALSE, warning=FALSE, eval=FALSE}
remotes::install_github("Greensway/BIRDS")
library(BIRDS)
```

```{r organisebirds, eval=FALSE}
OB <- organiseBirds(xf$data, sppCol = "species" , 
                    # We only want observations identified at the species level
                    taxonRankCol = "rank", taxonRank = "species", 
                    # the visits are defined by collector and named locality
                    idCols = c("locality", "collector"), 
                    timeCols = c("year", "month", "day"), 
                    xyCols = c("longitude","latitude") )

# We don't need the whole grid, just the piece that overlaps our searching polygon
wInt <- unlist(st_intersects(gotaland, Sweden_Grid_25km_Wgs84))
gotaland_grid25 <- Sweden_Grid_25km_Wgs84[wInt,]

SB <- summariseBirds(OB, grid = gotaland_grid25, spillOver = "unique")
```

```{r save_birds, eval=FALSE, include=FALSE}
save(list=c("OB", "SB"), file = "data/BIRDS_summary.rdata")
```
```{r load_birds, include=FALSE}
load(file = "data/BIRDS_summary.rdata")
```

Once summarised, we can see over space and for a few selected years how
the number of observations is distributed:

```{r plotBIRDSspatial_cose, eval=FALSE}
maxC <- max(SB$spatial$nObs, na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")
oldpar <- par()
par(mar = c(1,1,1,1), mfrow=c(1,3))
plot(SB$spatial$geometry, col=palBW(SB$spatial$nObs),
     border = "grey", main="All years") ## with palette
legend("bottomleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       title = "Number of \nobservations", pch = 15, bty="n")

## or export other combinations, e.g. one map per observed year
yearlySp <- exportBirds(SB, 
                        dimension = "spatial", 
                        timeRes = "yearly", 
                        variable = "nObs", 
                        method = "sum")

maxC <- max(yearlySp$'2005', na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")

plot(yearlySp$geometry, col=palBW(yearlySp$'2005'), 
     border = "grey",main="2005")
legend("bottomleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       border = "grey",
       title = "Number of \nobservations", pch = 15, bty="n")

maxC <- max(yearlySp'2010', na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")

plot(yearlySp$geometry, col=palBW(yearlySp$'2010'), 
     border = "grey",main="2010")
legend("bottomleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       border = "grey",
       title = "Number of \nobservations", pch = 15, bty="n")
par(oldpar)
```

```{r plotBIRDSspatial, echo=FALSE, fig.width=8}
maxC <- max(SB$spatial$nObs, na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")
oldpar <- par()
par(mar = c(1,1,1,1), mfrow = c(1,3))
plot(SB$spatial$geometry, col = palBW(SB$spatial$nObs),
     border = "grey", main = "All years") ## with palette
legend("bottomleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       title = "Number of \nobservations", pch = 15, bty = "n")

## or export other combinations, e.g. one map per observed year
yearlySp <- exportBirds(SB, 
                        dimension = "spatial", 
                        timeRes = "yearly", 
                        variable = "nObs", 
                        method = "sum")

maxC <- max(yearlySp$'2005', na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")

plot(yearlySp$geometry, col = palBW(yearlySp$'2005'), 
     border = "grey", main = "2005")
legend("bottomleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       border = "grey",
       title = "Number of \nobservations", pch = 15, bty = "n")

maxC <- max(yearlySp$'2010', na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")

plot(yearlySp$geometry, col = palBW(yearlySp$'2010'), 
     border = "grey",main = "2010")
legend("bottomleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       border = "grey",
       title = "Number of \nobservations", pch = 15, bty = "n")
suppressWarnings(par(oldpar))
```

We now want to use the number of field visits as the measure for
sampling effort. :

```{r ggplot1, message=FALSE, warning=FALSE, fig.width=8}
library(cowplot)
library(ggplot2)
library(colorRamps)
library(gridExtra)

vis <- ggplot(data = SB$spatial, aes( fill = nVis)) +
  geom_sf() +
  ggtitle("Visits") +
  scale_fill_gradient(low = "#56B1F7",
                      high = "#132B43",
                      na.value = NA) +
  theme(plot.margin = margin(1, 1, 1, 1, "pt")) +
  theme_cowplot()

spp <- ggplot(data = SB$spatial, aes( fill = nSpp)) +
  geom_sf() +
  ggtitle("Number of species") +
  scale_fill_gradient(low = "#56B1F7",
                      high = "#132B43",
                      na.value = NA) +
  theme(plot.margin = margin(1, 1, 1, 1, "pt")) +
  theme_cowplot()

grid.arrange(vis, spp, ncol = 2)
```

#### Temporal check

We see that `SB` contains an element called `SB$temporal` that contains
a daily time series with time-specific rows when there is information.
`xts` also supports day time, but dating below day resolution is not yet
implemented in the `BIRDS` package.

```{r temporal}
sb.xts <- SB$temporal
head(sb.xts, 5)
```

Sub-setting is convenient in `xts` as you can do it with its dates and
with a `/` for a range of dates.

```{r subsetting, message=FALSE, warning=FALSE}
sb.xts["2010-09-07"] #a specific day
sb.xts["2010-09-01/2010-09-15"] #for a period
sb.xts["2010-09"] #a specific month
```

The package `xts` has several tools for converting to different time
periods. Here we use `apply.monthly` to obtain the total number of
observations and visits per month. The plot command for an object of
calss `xts` provides a many features. This makes it fairly easy to
customize your plots. Read more in `?plot.xts`.

```{r monthlyPlotcode, eval=FALSE}
library(xts)
obs.m <- apply.monthly(sb.xts$nObs, "sum", na.rm = TRUE)
vis.m <- apply.monthly(sb.xts$nVis, "sum", na.rm = TRUE)

plot(obs.m, 
     col = "darkblue", 
     grid.ticks.on = "month", 
     major.ticks = "year", 
     grid.col = "lightgrey",  
     main = "Total number of daily observations and visits per month")

lines(vis.m, col = "orange", lwd = 2, on = 1)
```

```{r monthlyPlot, echo=FALSE, include=FALSE}
library(xts)
obs.m <- apply.monthly(sb.xts$nObs, "sum", na.rm = TRUE)
vis.m <- apply.monthly(sb.xts$nVis, "sum", na.rm = TRUE)

plot(obs.m, 
     col = "darkblue", 
     grid.ticks.on = "month", 
     major.ticks = "year", 
     grid.col = "lightgrey",  
     main = "Total number of daily observations and visits per month")

lines(vis.m, col = "orange", lwd = 2, on = 1)
```

```{r, echo=FALSE, fig.width=8}
addLegend("topleft", c("Observations", "Visits"), 
          col = c("darkblue","orange"), 
          lwd = 2)
```

## Species trends

We can now look at some particular species and ask whether those have
changed in occurrence over time:

```{r sppTrends, message=FALSE}
speciesSummary(SB)[,1:4]
```

We pick two species and compare their trends in number of visits where
the species where reported, relative to the total number of visits.

```{r trends, fig.width=8}
library(dplyr)
sppCount <- obsData(OB) |> 
    group_by(year, visitUID) |> 
    summarise("focalCountLq" = sum(scientificName == "Libellula quadrimaculata"),
              "focalCountSd" = sum(scientificName == "Sympetrum sanguineum"),
              "sppLength" = length(unique(scientificName)), 
              .groups = "drop") |> 
    ungroup() |> 
    group_by(year) |> 
    summarise("focalCountLq" = sum(focalCountLq),
              "focalCountSd" = sum(focalCountSd),
              "nVis" = length(unique(visitUID)),
              "relCountLq" = focalCountLq / nVis,
              "relCountSd" = focalCountSd / nVis,
              .groups = NULL)

oldpar <- par(no.readonly = TRUE)
plot(sppCount$year, sppCount$relCountLq, 
     type = "l", lwd = 3, xlab = "Year", 
     ylab = "Relative number of visits with observations", 
     ylim = c(0, max(sppCount$relCountLq)), 
     xaxp = c(2000, 2010, 10))
lines(sppCount$year, sppCount$relCountSd, lwd = 3, col = "#78D2EB")
legend("bottomright", 
       legend = c("Libellula quadrimaculata","Sympetrum sanguineum"), 
       text.font = 3, col = c("black", "#78D2EB"), lwd = 3, bty = "n")
par(oldpar)
```
