---
title: "Using R tools for analysis <br>of primary biodiversity data provided by SBDI"
author: "Debora Arlt and Alejandro Ruete <br>for the Swedish Biodiversity Data Infrastructure" 
date: "`r Sys.Date()`"
description: "R tools for biodiversity data" 
site: bookdown::bookdown_site

bibliography: references.bib
biblio-style: apalike
link-citations: yes
colorlinks: yes

fontsize: 10pt
github-repo: biodiversitydata-se/r-tools-tutorial
cover: images/sbdi-logo-orginal-large.png
favicon: images/favicon_io/favicon.ico
url: https\://biodiversitydata-se.github.io/r-tools-tutorial/
---

# Introduction {.unnumbered}

Placeholder


## R and Mirroreum{.unnumbered}
## SBDI4R - R package to search an access data{.unnumbered}
## Customizing SBDI4R{.unnumbered}
### Caching{.unnumbered}
### E-mail address{.unnumbered}
### Setting the download reason{.unnumbered}
### Other options{.unnumbered}
## Other packages needed {.unnumbered}
## Your collaboration is appreciated {.unnumbered}

<!--chapter:end:index.Rmd-->


# Example with fish data from SERS

Placeholder


## Plotting data on a map
#### Leaflet
## Temporal summary
## Species summary
## Spatial biodiversity analysis

<!--chapter:end:01-fish-data.Rmd-->

# Example with opportunistic data on Dragonflies

In this example we are interested in exploring opportunistically collected data 
from the Swedish citizen science observation data portal - Artportalen.

## Name searching
To begin, we want be sure there is an unequivocal way to find the species within 
the order Odonata and nothing else, so let's search for it:
```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE, include=FALSE}
library(SBDI4R)
```

```{r search1, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
sx <- search_fulltext("odonata")
sx$data[, c("guid", "scientificName", "rank", "occurrenceCount")]
```  

```{r search1eval, include=FALSE}
sx <- search_fulltext("odonata")
```  
```{r search1show, echo=FALSE}
sx$data[, c("guid", "scientificName", "rank", "occurrenceCount")]
```


We quickly see there that other taxonomic definitions appear too, but there is only one order. 
Let's refine the search. To know which search fields we can use to filter the search 
we use the function `sbdi_fields(fields_type = "general")`.
The search field we are looking for is "order_s".  
```{r search2, warning=FALSE, message=FALSE, error=FALSE}
sx <- search_fulltext(fq="order_s:Odonata", page_size = 10)
sx$data[, c("scientificName", "rank", "occurrenceCount")]
```  
```{r search2eval, include=FALSE}
sx <- search_fulltext("odonata")
```  
```{r search2show, echo=FALSE}
sx$data[, c("guid", "scientificName", "rank", "occurrenceCount")]
```

Now we can download the taxonomic data (note that the search is case-sensitive):
```{r, message=FALSE, results=FALSE}
tx <- taxinfo_download("order_s:Odonata", 
                       fields = c("guid", "order_s","genus_s", "specificEpithet_s", 
                                  "scientificName",  "canonicalName_s", "rank"), 
                       verbose = FALSE)
tx <- tx[tx$rank == "species" & tx$genusS != "",] ## restrict to species and not hybrids
```
Then you can save `tx` as the complete species list for later use. 

## Get the observations, filter the search, get quality assertions, plotting data on a map and save data
As usual we start by searching for the data resource we are interested in using 
the function `pick_filter()`. This is an interactive query guiding you through 
the many resources available to filtering your query (data resources, spatial 
layers, and curated species lists).

```{r filter2.1, message=FALSE, eval=FALSE}
# follow the instructions 
fq_str <- pick_filter("resource") 
``` 

Follow the instruction. Your choices here would have been "in3" --> "dr5". 
Your variable `fq_str` will now contain a string "data_resource_uid:dr5".

```{r filter2.2, echo=FALSE, eval=TRUE}
# we cant do it interactive here so we force it
fq_str <- "data_resource_uid:dr5"
```

We only need data from 2000 to 2010
```{r filter2.3}
y1 <- 2000
y2 <- 2010
fq_str <- c(fq_str, paste0("year:[", y1, " TO ", y2,"]"))
# Note the square brackets are hard limits
```

Select data – get records for Southern Sweden ([Götaland](https://en.wikipedia.org/wiki/G%C3%B6taland)).

Vector spatial layers (eg. polygons) can be imported  in a number of different ways. 
SBDI APIs take as search input polygons in the s.k. WKT [Well Known Text](https://www.geoapi.org/3.0/javadoc/org/opengis/referencing/doc-files/WKT.html) 
format. So the first step is to load a vector layer and transform it into a WKT string. 
You could instead use the data we kindly provided in the SBDI4R package `data("swe")`. 

```{r sweshape}
data("swe")
wGotaland <- swe$Counties$LnNamn %in% c("Blekinge", "Gotlands", "Hallands", "Jönköpings", "Kalmar", "Kronobergs", "Östergötlands", "Skåne", "Västra Götalands")
gotaland_c <- swe$Counties[wGotaland,]
```

We could create the WKT string using the `rgeos` library:
```{r, eval=FALSE}
library(rgeos)
wkt <- writeWKT(gotaland_c)
```

Unfortunately, in this instance this gives a WKT string that is too long and won't
be accepted by the web service. Also, the shapefile we just got is projected in 
the coordinate system SWEREF99 TM, and the web service only accepts coordinates in 
a geodesic coordinate system WGS84. Instead, let's construct the WKT string directly, 
which gives us a little more control over its format:
```{r, warning=FALSE}
gotaland_c <- sf::as_Spatial(
                sf::st_transform(
                  sf::st_as_sf(gotaland_c), 
                  crs = sf::st_crs(4326)$wkt) )

gotaland <- rgeos::gUnaryUnion(gotaland_c)

# extract the polygons coordinates
nPol <- length(gotaland@polygons[[1]]@Polygons) 
lonlat <- list()
for(p in seq(nPol)){
  lonlat[[p]] <- gotaland@polygons[[1]]@Polygons[[p]]@coords
}
lonlat <- do.call(rbind, lonlat)

# create a convex hull of the polygon to reduce the length of the WKT string
gotaland_ch <- chull(lonlat)
lonlat <- lonlat[c(gotaland_ch, gotaland_ch[1]), ]

# create WKT string
# first join each lon-lat coordinate pair
wkt_temp <- apply(lonlat, 1, function(z) paste(round(z,4), collapse=" "))
# now build the WKT string
wkt <- paste("MULTIPOLYGON(((", paste(wkt_temp, collapse=","), ")))", sep="")
# NOTE: as of today, the SBDI APIs will only work properly if the polygon is 
# submitted as a MULTIPOLYGON
```


```{r getData}
sbdi_fields("occurrence")[,1:2]
xf <- SBDI4R::occurrences(taxon = "order:Odonata", 
                  fq = fq_str,
                  wkt = wkt,
                  extra = "collector",
                  email = "sbdi4r-test@biodiversitydata.se", 
                  download_reason_id = 10)

xf$meta
```

but before we can use the observation records we need to know how the observation effort has varied over time and in space. For this we define field visits i.e. occasions at which an observer has sampled observations – if we have information on observer id, location id and date we can aggregate data into “field visits”. We do this using BIRDS, and 25km grid:

```{r birds}
library(BIRDS)
OB <- organiseBirds(xf$data, sppCol = "species" , 
                    taxonRankCol = "rank", taxonRank = "species", 
                    idCols = c("locality", "collector"), 
                    timeCols = c("year", "month", "day"), 
                    xyCols =c("longitude","latitude") )

gotaland_grid25 <- raster::intersect(gotaland, Sweden_Grid_25km_Wgs84)

# gotaland_grid25 <- gIntersection(spTransform(Sweden_Grid_25km_Wgs84, 
#                                              CRSobj = CRS(sf::st_crs(4326)$wkt)),
#                                  gotaland,
#                                  byid = TRUE)

SB <- summariseBirds(OB, grid = gotaland_grid25, spillOver = "unique")
```

```{r plotBIRDSspatial, message=FALSE, warning=FALSE}
maxC <- max(SB$spatial@data$nObs, na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")
oldpar <- par()
par(mar = c(4,0,4,0), mfrow=c(1,3))
plot(SB$spatial, col=palBW(SB$spatial@data$nObs),
     border = "grey", main="All years") ## with palette
legend("topleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       title = "Number of \nobservations", pch = 15, bty="n")

## or export other combinations, e.g. one map per observed year
yearlySp <- exportBirds(SB, 
                        dimension = "spatial", 
                        timeRes = "yearly", 
                        variable = "nObs", 
                        method = "sum")

maxC <- max(yearlySp@data$'2005', na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")

plot(yearlySp["2005"], col=palBW(yearlySp@data$'2005'), 
     border = "grey",main="2005")
legend("topleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       border = "grey",
       title = "Number of \nobservations", pch = 15, bty="n")

maxC <- max(yearlySp@data$'2010', na.rm = TRUE)
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, maxC), 
                               na.color = "transparent")

plot(yearlySp["2010"], col=palBW(yearlySp@data$'2010'), 
     border = "grey",main="2010")
legend("topleft", inset = c(0,0.05),
       legend = round(seq(0, maxC, length.out = 5)),
       col = palBW(seq(0, maxC, length.out = 5)),
       border = "grey",
       title = "Number of \nobservations", pch = 15, bty="n")
suppressWarnings(par(oldpar))

```
```{r ggplot1, fig.width=5}
library(sf)
library(cowplot)
library(ggplot2)
library(colorRamps)

spatial_sf <- st_as_sf(SB$spatial)

obs <- ggplot(data = spatial_sf, aes( fill = nVis))+
  geom_sf()+
  ggtitle("Visits")+
  scale_fill_gradient(low = "#56B1F7",
                      high = "#132B43",
                      na.value = NA) +
  theme_cowplot()

obs
```

```{r ggplot2, fig.width=5}
spp <- ggplot(data = spatial_sf ,aes( fill = nSpp))+
  geom_sf()+
  ggtitle("Number of species")+
  scale_fill_gradient(low = "#56B1F7",
                      high = "#132B43",
                      na.value = NA) +
  theme_cowplot()

spp
```

How has observation effort (frequency of visits) varied over time and space? – 1) show maps as in Example 7 (all years, year 2000, 2002, 2004, 2006, 2008, 2010), 2 make also a time line plot with no. visits against years, no. of gridcells with visits against years.

we see that SB contains an element called `SB$temporal` that contains a daily time series with time specific rows when there is information. `xts` also supports time, but dating below day resolution is not yet implemented in the `BIRDS` package.  

```{r temporal}
sb.xts <- SB$temporal
head(sb.xts)
dim(sb.xts)
```


Sub-setting is convenient in `xts` as you can do it with its dates and with a `/` for a range of dates.  

```{r subsetting, echo=TRUE, message=FALSE, warning=FALSE}
sb.xts["2010-09"] #a specific month
sb.xts["2010-09-07"] #a specific day
sb.xts["2007-01-01/2007-05-01"] #for a period
```

The package `xts` has several tools for converting to different periods. Here we will use `to.monthly`. This provides, the first, min, max, and last of the data. We can plot the daily maximum number of observations. The plot command with an `xts` object provides a TON of features. This makes it fairly easy to customize your plots. Read more in `?plot.xts`.  

```{r monthly and plot, echo=TRUE, fig.height=5, fig.width=9, message=FALSE, warning=FALSE}
library(xts)
obs.m <- to.monthly(sb.xts$nObs) 
obs.m["2007-04"]
sb.xts["2007-04"]
plot(obs.m["2000/2010",2], col = "darkblue", grid.ticks.on = "month", 
     major.ticks = "month", grid.col = "lightgrey",  
     main = "Maximum number of daily observations per month")

vis.m <- to.monthly(sb.xts$nVis) 
vis.m["2007-04"]
sb.xts["2007-04"]
plot(vis.m["2000/2010",2], col = "darkblue", grid.ticks.on = "month", 
     major.ticks = "month", grid.col = "lightgrey", 
     main = "Maximum number of daily visits per month")
```


We can now look at some particular species and ask whether this has changed in occurrence over time:
Plot no. records of species x and no. visits all species over years (we simply explore by comparing records for a species with no visits, can assume that species has increased of stronger positive trend than for no. visits)

Plot no. gridcells with visits for species x and no. gridcells with visits for all species over years (we simply explore by comparing records for a species with no visits, can assume that species has increased of stronger positive trend than for no. visits)
(species x: Tvåfläckad trollslända Epitheca bimaculata)


<!--chapter:end:02-dragonflies-data.Rmd-->

